|

### 2023-05-04 paper.Rmd
final changes

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the availability of many numerical hypothesis test procedures designed to use residuals to assess problems with a model fit. Here we provide evidence for why this is good advice using data from a visual inference experiment. We show how conventional tests are too sensitive, which means that too often the conclusion would be that the model fit is inadequate. The experiment uses the lineup protocol which puts a residual plot in the context of null plots. This helps generate reliable and consistent reading of residual plots for better model diagnosis. It can also help in an obverse situation where a conventional test would fail to detect a problem with a model due to contaminated data. The lineup protocol also serves as an omnibus test, capable of detecting a range of departures from good residuals simultaneously.
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

Table \ref{tab:example-residual-plot-table} displays the $p$-values from the RESET, BP and SW tests applied to the residual plots in Figure \ref{fig:residual-plot-common-departures}. The RESET test and BP test were computed using the `resettest` and `bptest` functions from the R package `lmtest`, respectively. The SW test was computed using the `shapiro.test` from the core R package `stats`. ^[Although we didn't use it, it's useful to know that the R package `skedastic` [@skedastic] also contains a large collection of functions to test for heteroskedasticity.] Although, the RESET test is exact, it requires the selection of a power parameter.  @ramsey1969tests recommends a power of four, which we adopted in our analysis.  The BP and SW tests are approximate.

@@

As pointed out by @vanderplas2021statistical, this basic binomial model is deficient. It does not take into account the possible dependencies in the visual test due to repeated evaluations of the same lineup, or account for when participants are offered the option to select one or more "most different" plots, or none, from a lineup. They suggest three common lineup scenarios: (1) $K$ different lineups are shown to $K$ participants, (2) $K$ lineups with different null plots but the same data plot are shown to $K$ participants, and (3) the same lineup is shown to $K$ participants. Scenario 3 is the most feasible to apply, but has the most dependencies to accommodate for the $p$-value calculation. For Scenario 3, VanderPlas et al propose modelling the probability of plot $i$ being selected from a lineup as $\theta_i$, where $\theta_i \sim Dirichlet(\alpha)$ for $i=1,...,m$ and $\alpha > 0$. The number of times plot $i$ being selected in $K$ evaluations is denoted as $c_i$. In case participant $j$ makes multiple selections, $1/s_j$ will be added to $c_i$ instead of one, where $s_j$ is the number of plots participant $j$ selected for $j=1,...K$. This ensures $\sum_{i}c_i=K$. Since we are only interested in the selections of the data plot $i$, the marginal model can be simplified to a beta-binomial model and thus the visual $p$-value is given as

@@

The parameter $\alpha$ used in Equation \ref{eq:pvalue-beta-binomial} and \ref{eq:pmf-beta-binomial} is usually unknown and will need to be estimated from the survey data. An interpretation of $\alpha$ is that when it is low only a few plots are attractive to the observers and tend to be selected, and when high, most plots are equally likely to be chosen. VanderPlas et al define $c$-interesting plot to be if $c$ or more participants select the plot as the most different. The expected number of plots selected at least $c$ times, $E[Z_c]$, is then calculated as

```{=tex}
\begin{equation} \label{eq:c-interesting-expectation}
E[Z_c(\alpha)] = \frac{m}{B(\alpha, (m-1)\alpha)}\sum_{\lceil c \rceil}^{K}{K \choose x} B(x + \alpha, K - x + (m-1)\alpha).\end{equation}
```

@@

Since $supp(X) = [-1, 1]$, choosing $a$ to be $-1$, $0$ and $1$ can generate "left-triangle", "butterfly" and "right-triangle" shapes as displayed in Figure \ref{fig:different-shape-of-heter}. The term $(2 - |a|)$ maintains the magnitude of residuals across different values of $a$. Figure \ref{fig:different-b} shows the butterfly shape as the ratio parameter ($b$) is changed.

@@

a_labels <- c("left-triangle", "butterfly", "right-triangle")

@@

- uniform, $U(-1, 1)$, 
- normal, $N(0, 0.3^2)$, 
- skewed, $lognormal(0, 0.6^2)/3$, and 
- discrete, $U\{1, 5\}$.

@@

\begin{equation}
E = \frac{1}{2}\left(\boldsymbol{\mu}_z'(diag(\boldsymbol{R}\sigma^2))^{-1}\boldsymbol{\mu}_z\right)
\label{eq:nonlin-effect-size}
\end{equation}

\noindent where $diag(.)$ is the diagonal matrix constructed from the diagonal elements of a matrix,
$\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{H}$ is the residual operator, $\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'$ is the hat matrix, $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of residuals where $\boldsymbol{Z}$ contain any higher order terms of $\boldsymbol{X}$ left out of the regression equation, $\boldsymbol{\beta}_z$ contains the corresponding coefficients, and $\sigma^2\boldsymbol{I}$ is the assumed covariance matrix of the error term when $H_0$ is true.

@@

\begin{equation}
E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')|}{|diag(\boldsymbol{R})|} - n + tr(diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')^{-1}diag(\boldsymbol{R}))\right)
\label{eq:hetero-effect-size}
\end{equation}

@@

Data collection used a total of `r filter(vi_survey, !attention_check) %>% count(unique_lineup_id) %>% nrow()` lineups, and resulted in a total of `r nrow(filter(vi_survey, !attention_check))` evaluations from `r vi_survey %>% count(set, exp) %>% nrow()` participants. Roughly half corresponded to the two models, non-linearity and heteroskedasticiy, and the three collection periods had similar numbers of evaluations. Each participant received two of the `r vi_survey %>% filter(attention_check) %>% count(unique_lineup_id) %>% nrow()` attention check lineups which were used to filter results of participants who were clearly not making an honest effort (only 11 of `r vi_survey %>% count(set, exp) %>% nrow() + 11`). To estimate $\alpha$ for calculating statistical significance (see Section \ref{sig}) there were `r nrow(filter(vi_survey, null_lineup, !attention_check))` evaluations of `r vi_survey %>% filter(null_lineup, !attention_check) %>% count(unique_lineup_id) %>% nrow()` null lineups. Neither the attention checks nor null lineups were used in the subsequent analysis. The de-identified data, `vi_survey`, is made available in the R package, `visage`.^[The R package `visage` is currently available on GitHub (https://github.com/TengMCing/visage).]

@@

  # Draw boot visual test curves
  geom_line(data = pred_visual_poly_boot,
            aes(log_effect_size, power, col = "Visual test", group = boot_id),
            size = 1,
            alpha = 0.01) +

  # Draw visual test power curve
  geom_line(data = pred_visual_poly,
            aes(log_effect_size, power, col = "Visual test"), 
            size = 1) +

@@

  # Draw boot visual test curves
  geom_line(data = pred_visual_heter_boot,
            aes(log_effect_size, power, col = "Visual test", group = boot_id),
            size = 1,
            alpha = 0.01) +

  # Draw visual test power curve
  geom_line(data = pred_visual_heter,
            aes(log_effect_size, power, col = "Visual test"), 
            size = 1) +

@@

# Mosaic plot
p_value_cmp_dat %>%
  ggplot() +
  geom_mosaic(aes(x = ggmosaic::product(reject, conv_reject), 
                  fill = reject)) +
  facet_grid(~type) +
  ylab("Visual tests") +
  xlab("Conventional tests") +
  labs(fill = "Conventional tests") +
  scale_fill_brewer("", palette = "Dark2") +
  theme_bw() +
  theme(legend.position = "none") +
  coord_fixed()
```

@@

In regression analysis, predictions are conditional on the observed values of the predictors, that is, the conditional mean of the dependent variable $Y$ given the value of the independent variable $X$, $E(Y|X)$. This is an often forgotten element of regression analysis but it is important. Where $X$ is observed, the distribution of the $X$ values in the sample, or consequently $\hat{Y}$, may affect the ability to read any patterns in the residual plots. The effect of fitted value distribution on test performance is assess using four different distributions of fitted values: uniform, normal, discrete and lognormal (skewed). We expect that if $\hat{Y}$ has a uniform distribution, it is easier to read the relationship with the residuals.

@@

dat_uniform_5_pred <- dat_uniform_5 %>%
  nest(dat = c(effect_size, offset0, reject)) %>%
  mutate(mod = map(dat, 
                   ~glm(reject ~ effect_size - 1, 
                        family = binomial(), 
                        data = .x,
                        offset = offset0))) %>%
  mutate(power = map2(mod, type, function(mod, type) {
    this_min_es <- ifelse(type == "heteroskedasticity", min_heter_es, min_poly_es)
    this_max_es <- ifelse(type == "heteroskedasticity", max_heter_es, max_poly_es)
    result <- data.frame(effect_size = seq(this_min_es[1], this_max_es[1], 0.1),
                         offset0 = log(0.05/0.95))
    result$power <- predict(mod, type = "response", newdata = result)
    result
  })) %>%
  select(-dat, -mod) %>%
  unnest(power) %>%
  mutate(log_effect_size = log(effect_size)) %>%
  mutate(type = factor(stringr::str_to_title(type), 
                       levels = c("Non-Linearity", "Heteroskedasticity")))
```

@@

diff_x_dist_power_pred %>%
  mutate(x_dist = ifelse(x_dist == "even_discrete", 
                         "Discrete", 
                         stringr::str_to_title(x_dist))) %>%
  mutate(type = factor(stringr::str_to_title(type), 
                       levels = c("Non-Linearity", 
                                  "Heteroskedasticity"))) %>%
  mutate(x_dist = factor(x_dist, 
                         levels = c("Uniform", 
                                    "Normal", 
                                    "Lognormal", 
                                    "Discrete"))) %>%

@@

  # Draw boot uniform (5 eval) lines
  geom_line(data = dat_uniform_5_pred,
            aes(log_effect_size, power, col = "Uniform", group = boot_id),
            size = 1,
            alpha = 0.01) +

@@

>>>>

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the availability of many numerical hypothesis test procedures designed to use residuals to assess problems with a model fit. Here we provide evidence for why this is good advice using data from a visual inference experiment. We show how conventional tests are too sensitive, which means that too often the conclusion would be that the model fit is inadequate. The experiment uses the lineup protocol which puts a residual plot in the context of null plots. This helps generate reliable and consistent reading of residual plots for better model diagnosis. It can also help in an obverse situation where a conventional test would fail to detect a problem with a model due to contaminated data. The lineup protocol also detects a range of departures from good residuals simultaneously.
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

Table \ref{tab:example-residual-plot-table} displays the $p$-values from the RESET, BP and SW tests applied to the residual plots in Figure \ref{fig:residual-plot-common-departures}. The RESET test and BP test were computed using the `resettest` and `bptest` functions from the R package `lmtest`, respectively. The SW test was computed using the `shapiro.test` from the core R package `stats`. ^[Although we didn't use it, it's useful to know that the R package `skedastic` [@skedastic] also contains a large collection of functions to test for heteroskedasticity.] The RESET test requires the selection of a power parameter.  @ramsey1969tests recommends a power of four, which we adopted in our analysis.

@@

As pointed out by @vanderplas2021statistical, this basic binomial model is deficient. It does not take into account the possible dependencies in the visual test due to repeated evaluations of the same lineup, or account for when participants are offered the option to select one or more "most different" plots, or none, from a lineup. They suggest three common lineup scenarios: (1) $K$ different lineups are shown to $K$ participants, (2) $K$ lineups with different null plots but the same data plot are shown to $K$ participants, and (3) the same lineup is shown to $K$ participants. Scenario 3 is the most feasible to apply, but has the most dependencies to accommodate for the $p$-value calculation. For Scenario 3, VanderPlas et al propose modelling the probability of plot $i$ being selected from a lineup as $\theta_i$, where $\theta_i \sim \text{Dirichlet}(\alpha)$ for $i=1,...,m$ and $\alpha > 0$. The number of times plot $i$ being selected in $K$ evaluations is denoted as $c_i$. In case participant $j$ makes multiple selections, $1/s_j$ will be added to $c_i$ instead of one, where $s_j$ is the number of plots participant $j$ selected for $j=1,...K$. This ensures $\sum_{i}c_i=K$. Since we are only interested in the selections of the data plot $i$, the marginal model can be simplified to a beta-binomial model and thus the visual $p$-value is given as

@@

The parameter $\alpha$ used in Equation \ref{eq:pvalue-beta-binomial} and \ref{eq:pmf-beta-binomial} is usually unknown and will need to be estimated from the survey data. An interpretation of $\alpha$ is that when it is low only a few plots are attractive to the observers and tend to be selected, and when high, most plots are equally likely to be chosen. VanderPlas et al define $c$-interesting plot to be if $c$ or more participants select the plot as the most different. The expected number of plots selected at least $c$ times, $\mathrm{E}[Z_c]$, is then calculated as

```{=tex}
\begin{equation} \label{eq:c-interesting-expectation}
\mathrm{E}[Z_c(\alpha)] = \frac{m}{B(\alpha, (m-1)\alpha)}\sum_{\lceil c \rceil}^{K}{K \choose x} B(x + \alpha, K - x + (m-1)\alpha).\end{equation}
```

@@

Since $\text{supp}(X) = [-1, 1]$, choosing $a$ to be $-1$, $0$ and $1$ can generate "left-triangle", "butterfly" and "right-triangle" shapes as displayed in Figure \ref{fig:different-shape-of-heter}. The term $(2 - |a|)$ maintains the magnitude of residuals across different values of $a$. Figure \ref{fig:different-b} shows the butterfly shape as the ratio parameter ($b$) is changed.

@@

a_labels <- c("Left-triangle", "Butterfly", "Right-triangle")

@@

- uniform, $U(-1, 1)$, 
- normal, $N(0, 0.3^2)$, 
- skewed, $\text{lognormal}(0, 0.6^2)/3$, and 
- discrete, $U\{1, 5\}$.

@@

\begin{equation}
E = \frac{1}{2}\left(\boldsymbol{\mu}_z'(\text{diag}(\boldsymbol{R}\sigma^2))^{-1}\boldsymbol{\mu}_z\right)
\label{eq:nonlin-effect-size}
\end{equation}

\noindent where $\text{diag}(.)$ is the diagonal matrix constructed from the diagonal elements of a matrix,
$\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{H}$ is the residual operator, $\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'$ is the hat matrix, $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of residuals where $\boldsymbol{Z}$ contain any higher order terms of $\boldsymbol{X}$ left out of the regression equation, $\boldsymbol{\beta}_z$ contains the corresponding coefficients, and $\sigma^2\boldsymbol{I}$ is the assumed covariance matrix of the error term when $H_0$ is true.

@@

\begin{equation}
E = \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')|}{|\text{diag}(\boldsymbol{R})|} - n + \text{tr}(\text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')^{-1}\text{diag}(\boldsymbol{R}))\right)
\label{eq:hetero-effect-size}
\end{equation}

@@

Data collection used a total of `r filter(vi_survey, !attention_check) %>% count(unique_lineup_id) %>% nrow()` lineups, and resulted in a total of `r nrow(filter(vi_survey, !attention_check))` evaluations from `r vi_survey %>% count(set, exp) %>% nrow()` participants. Roughly half corresponded to the two models, non-linearity and heteroskedasticiy, and the three collection periods had similar numbers of evaluations. Each participant received two of the `r vi_survey %>% filter(attention_check) %>% count(unique_lineup_id) %>% nrow()` attention check lineups which were used to filter results of participants who were clearly not making an honest effort (only 11 of `r vi_survey %>% count(set, exp) %>% nrow() + 11`). To estimate $\alpha$ for calculating statistical significance (see Section \ref{sig}) there were `r nrow(filter(vi_survey, null_lineup, !attention_check))` evaluations of `r vi_survey %>% filter(null_lineup, !attention_check) %>% count(unique_lineup_id) %>% nrow()` null lineups. Neither the attention checks nor null lineups were used in the subsequent analysis. The de-identified data, `vi_survey`, is made available in the R package, `visage`.

@@

  # Draw boot visual test curves
  geom_line(data = pred_visual_poly_boot,
            aes(log_effect_size, power, col = "visual test", group = boot_id),
            size = 1,
            alpha = 0.01) +

  # Draw visual test power curve
  geom_line(data = pred_visual_poly,
            aes(log_effect_size, power, col = "visual test"), 
            size = 1) +

@@

  # Draw boot visual test curves
  geom_line(data = pred_visual_heter_boot,
            aes(log_effect_size, power, col = "visual test", group = boot_id),
            size = 1,
            alpha = 0.01) +

  # Draw visual test power curve
  geom_line(data = pred_visual_heter,
            aes(log_effect_size, power, col = "visual test"), 
            size = 1) +

@@

# Mosaic plot
p_value_cmp_dat %>%
  mutate(type = ifelse(type == "non-linearity", "Non-linearity", "Heteroskedasticity")) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity"))) %>%
  ggplot() +
  geom_mosaic(aes(x = ggmosaic::product(reject, conv_reject), 
                  fill = reject)) +
  facet_grid(~type) +
  ylab("Visual tests") +
  xlab("Conventional tests") +
  labs(fill = "Conventional tests") +
  scale_fill_brewer("", palette = "Dark2") +
  theme_bw() +
  theme(legend.position = "none") +
  coord_fixed()
```

@@

In regression analysis, predictions are conditional on the observed values of the predictors, that is, the conditional mean of the dependent variable $Y$ given the value of the independent variable $X$, $\text{E}(Y|X)$. This is an often forgotten element of regression analysis but it is important. Where $X$ is observed, the distribution of the $X$ values in the sample, or consequently $\hat{Y}$, may affect the ability to read any patterns in the residual plots. The effect of fitted value distribution on test performance is assess using four different distributions of fitted values: uniform, normal, discrete and lognormal (skewed). We expect that if $\hat{Y}$ has a uniform distribution, it is easier to read the relationship with the residuals.

@@

dat_uniform_5_pred <- dat_uniform_5 %>%
  nest(dat = c(effect_size, offset0, reject)) %>%
  mutate(mod = map(dat, 
                   ~glm(reject ~ effect_size - 1, 
                        family = binomial(), 
                        data = .x,
                        offset = offset0))) %>%
  mutate(power = map2(mod, type, function(mod, type) {
    this_min_es <- ifelse(type == "heteroskedasticity", min_heter_es, min_poly_es)
    this_max_es <- ifelse(type == "heteroskedasticity", max_heter_es, max_poly_es)
    result <- data.frame(effect_size = seq(this_min_es[1], this_max_es[1], 0.1),
                         offset0 = log(0.05/0.95))
    result$power <- predict(mod, type = "response", newdata = result)
    result
  })) %>%
  select(-dat, -mod) %>%
  unnest(power) %>%
  mutate(log_effect_size = log(effect_size)) %>%
  mutate(type = factor(ifelse(type == "non-linearity", "Non-linearity", "Heteroskedasticity"), 
                       levels = c("Non-linearity", "Heteroskedasticity")))
```

@@

diff_x_dist_power_pred %>%
  mutate(x_dist = ifelse(x_dist == "even_discrete", 
                         "discrete", 
                         x_dist)) %>%
  mutate(type = factor(ifelse(type == "non-linearity", "Non-linearity", "Heteroskedasticity"), 
                       levels = c("Non-linearity", "Heteroskedasticity"))) %>%
  mutate(x_dist = factor(x_dist, 
                         levels = c("uniform", 
                                    "normal", 
                                    "lognormal", 
                                    "discrete"))) %>%

@@

  # Draw boot uniform (5 eval) lines
  geom_line(data = dat_uniform_5_pred,
            aes(log_effect_size, power, col = "uniform", group = boot_id),
            size = 1,
            alpha = 0.01) +

@@

The R package `visage` is currently available at https://github.com/TengMCing/visage.
